---
title: "ModularityHW_Kelly"
author: "Michelle Kelly"
date: "June 4, 2018"
output: 
  pdf_document:
    df_print: paged
    fig_caption: yes
---  
<!-- MODULARITY LESSON 8: have defined purpose for each Rmd chunk-->
```{r setup, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
# Global options
knitr::opts_chunk$set(fig.height = 5, fig.width = 15, cache=TRUE) 
# Enable caching in all chunks, Set default size of figures to 5x9

# Packages
library(dplyr)
library(data.table)
library(ggplot2)
library(maps)
library(maptools)
library(mapdata)
library(viridis)
library(cowplot)
library(grid)
library(gridExtra)
```

```{r dataImport, echo=FALSE}
# Dependencies: 
#     USAAnnualPcpn1950_2008.rds
#     USAAnnualTemp1950_2008.rds

provided <- TRUE # TRUE if data loaded from file, FALSE if data pulled from API

if (provided == TRUE){
  precip <- readRDS(file = "USAAnnualPcpn1950_2008.rds") # annual precipitation data for USA
  temp <- readRDS(file = "USAAnnualTemp1950_2008.rds") # annual temperature data for USA
}
```

# Abstract  

(you will have to write this) climate change, long term trends, etc  

# Introduction  

  Average surface temperature on Earth has increased by 0.7&deg;C since the onset of the industrial era (UNFCC report 2007). Widespread fossil fuel use, and the following spike in atmospheric CO2, have acted as a global insulation blanket: trapping heat in the atmosphere and slowing it's release (Asfaw et al 2017). Increased temperatures are also impacting precipitation regimes; in the United States, increases in the severity of drought and flood events are expected (Dore, 2005). However, the impact of global change on a local level is expected to be more spatially heterogenous, with different regions of the US experiencing varied climate and precipitation responses (Trenberth 2011).  

  We assess these expectations using 50 years of weather station data, asking (1) can climate warming be seen across the United States, (2) are the effects of climate warming regionally-dependent, and (3) what change has occured in precipitation?

# Methods  

  We used annualy averaged weather-station temperature (`r length(unique(temp$state))` states, `r length(unique(temp$name))` weather stations) and precipitation data (`r length(unique(precip$state))` states, `r length(unique(precip$name))` weather stations) from 1950 to 2008. Only weather stations with at least 40 years of measurement data were included in this analysis.

  Yearly summary statistics were calculated by taking the mean observation within a state during any given year. To compare changes across time, a mean value for the duration of observations within a state was calculated, and subtracted from the annual value \ref{eq:1}. Therefore, negative values represent time periods when a state was experiencing below average temperatures, and positive values represent periods when a state was experiencing above average temperatures.  

$\label{eq:1} Deviation from average_{state} = Annual average_{state} - Time series average_{state} $

# Results  

```{r dataCleanup, echo=FALSE}
# MODULARITY LESSON 1: variables, not constants
minYears <- 40 # minimum years of data for each station

# MODULARITY LESSON 7: Pseudocode for trimming function

# trim <- function(dataset, minimumTimespan){
#   exclude years that have no data
#   count number of years of data for each station
#   if (years of data <= 40){
#     exclude station from dataframe
#   }
#   return(dataframe)
# }

# MODULARITY LESSON 2: write extensible code by creating trimming function

cleanByYear <- function(dataframe, nYears){
  dataframe <- na.exclude(dataframe)
  dataframe$count <- table(dataframe$name)[dataframe$name]
  for (row in 1:nrow(dataframe)){
    if (dataframe$count[row] <= nYears){
      dataframe[row,] <- NA
    }
  }
  dataframe <- na.exclude(dataframe)
  dataframe$count <- NULL
  return(dataframe)
}

precip <- cleanByYear(precip, minYears)
temp <- cleanByYear(temp, minYears)
```

```{r dataAnalysis, echo = FALSE}
# yearlyStatistics: mean data across USA for each year
yearlyStatistics <- function(dataframe){
  dataframe %>% group_by(year) %>% summarize(mean = mean(data))
}

precip_summary <- yearlyStatistics(precip)
temp_summary <- yearlyStatistics(temp)

# stateStatistics: mean data across each state for each year

# stateStatistics pseudocode:
#   1. compute overall average, so we can compare values to the average
#   2. group the data by year and state
#   3. compute the mean, standard deviation, and difference of values from the yearly average
#       where negative difference values represent colder/less rain than average
#       positive values represent warmer/more rain than average

stateStatistics <- function(dataframe){
  dataframe_state <- 
    dataframe %>% 
      group_by(year, state) %>% 
      summarize(mean = mean(data))
  
  dataframe_statesummary <- 
    dataframe_state %>%
      group_by(state) %>%
      summarise(state_mean = mean(mean))
  
  dataframe_merge <- 
    merge(dataframe_state, dataframe_statesummary, by = "state")
}

precip_state <- stateStatistics(precip)
temp_state <- stateStatistics(temp)

# deviation: compute difference between state yearly average and state overall (1950-2008) average
deviation <- function(dataset_state){
  dataset_state$mean - dataset_state$state_mean
}

precip_state$change <- deviation(precip_state)
temp_state$change <- deviation(temp_state)

# timeseries.change: compute mean deviation from state overall average across 1950-2008
timeseries.change <- function(dataset_state){
  dataset_state %>%
    group_by(state) %>%
    summarize(sum.change = sum(change))
}

precip.timeseries.change <- timeseries.change(precip_state)
temp.timeseries.change <- timeseries.change(temp_state)
```


```{r dataVis_overallPlot, echo=FALSE, include=FALSE}
# yearlyPlot: simple plot of yearly means, with added trendline
yearlyPlot <- function(dataset, ylabel, ...){
  ggplot(data = dataset, aes(x = year, y = mean, ...)) +
  geom_line() + geom_point() +
  geom_smooth(method = "lm") +
  xlab("Year") + ylab(ylabel) +
  theme_classic()
}

yearlyPlot(precip_summary, ylabel = "Yearly averaged precipitation (in)")
yearlyPlot(temp_summary, ylabel = "Yearly averaged temperature (F)")

```

```{r dataVis_functions, echo = FALSE}
# heatmap: visualization of change through time using colored tiles
heatmap <- function(dataset, legendlabel, ...){
  ggplot(dataset, aes(x = state, y = year, ...)) +
  geom_tile(aes(fill = change), color = "white") +
  scale_fill_distiller(type = "div", palette = 7, name = legendlabel) +
  xlab("State") + ylab("Year") +
  theme_classic() +
  theme(legend.position = "bottom")
}

# change.barplot: visualization of change through time using bars
change.barplot <- function(dataset_state, ylabel, col_number, ...){
  ggplot(dataset_state, aes(x = year, y = change, ...)) +
    geom_col(aes(fill = change)) +
    facet_wrap(~ state, ncol = col_number) +
    scale_fill_distiller(type = "div", palette = 7) +
    xlab("Year") + ylab(ylabel) +
    theme_classic() +
    theme(panel.background = element_rect(color = "black"),
          legend.position = "none")
}
```

```{r heatmap_temp, echo=FALSE, fig.width = 12, fig.height = 5, fig.cap="\\label{fig:figs}Anomalies in annual mean temperature, relative to 1950-2008 state average"}

# Anomalies in annual mean temperature, relative to 1950-2008 state average
heatmap(temp_state, legendlabel = "Temperature change (°F)")
```

```{r heatmap_precip, echo=FALSE, fig.width = 15, fig.height = 5, fig.cap="\\label{fig:figs}Anomalies in annual precipitation, relative to 1950-2008 state average"}

# Anomalies in annual precipitation, relative to 1950-2008 state average
grid.arrange(
  heatmap(precip_state, legendlabel = "Precipitation 
change (in)"),
  change.barplot(precip_state, ylabel = "Precipitation change (in)", col_number = 3),
  layout_matrix = rbind(c(1,2,2,2),
                        c(1,2,2,2))
)
```

```{r mapping, echo = FALSE, fig.width = 11, fig.height = 5, fig.cap="\\label{fig:figs}Average change in temperature over most recent decade (1998-2008), relative to 1950-2008 state average. "}
# data for mapping: plot mean deviation from average over last century of data (1998-2008) compared to mean deviation from average during 1950-1960
temp.change.map <- function(dataset_state, minYear, maxYear, axis_lims, legendPos){
  dataset_state <- 
    dataset_state %>%
      group_by(state, year) %>%
      filter(year >= minYear & year <= maxYear) %>%
      group_by(state) %>%
      summarize(mean.change = mean(change)) 
  
  states <- map_data("state") # pull state data from maps package
  states$region <- toupper(states$region) # change character data to uppercase
  setnames(states, "region", "state") # change column name
  
  dataset_state$state <- toupper(state.name[match(dataset_state$state, state.abb)])
  
  dataset_state <- inner_join(states, dataset_state, by = "state") # join dataframes
  
  # output map
  ggplot(data = dataset_state) +
    geom_polygon(aes(x = long, y = lat, group = group, fill = mean.change), 
                 color = "white") +
    coord_fixed(1.3) +
    scale_fill_distiller(limits = axis_lims, type = "div", palette = 7,
                         name = "Temperature 
change
relative to
average (°F)
") +
    theme_classic() +
    ggtitle(paste(minYear, maxYear, sep = "-")) +
    theme(legend.position = legendPos,
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
}

grid.arrange(
  temp.change.map(temp_state, minYear = 1950, maxYear = 1960, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 1960, maxYear = 1970, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 1970, maxYear = 1980, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 1980, maxYear = 1990, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 1990, maxYear = 2000, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 2000, maxYear = 2008, 
                  axis_lims = c(-2,2), legendPos = "none"),
  get_legend(temp.change.map(temp_state, minYear = 1950, maxYear = 1960, 
                             axis_lims = c(-2,2), legendPos = "right")),
  layout_matrix = rbind(c(1,2,3,7),
                        c(4,5,6,7)),
  widths = c(3,3,3,1), heights = c(2,2)
)

```

```{r significance_testing, echo = FALSE}
# significance testing: yearly means
sigtest_precip <- lm(precip_summary$mean ~ precip_summary$year)
summary(sigtest_precip) # not signif

sigtest_temp <- lm(temp_summary$mean ~ temp_summary$year)
summary(sigtest_temp) # not signif

# significance testing: change by state
summary(lm(temp_state$change ~ temp_state$year)) # signif

summary(lm(precip_state$change ~ precip_state$year)) #signif
```

# Discussion

# References

# Supplemental

```{r barplot_temp, echo = FALSE, fig.width = 17, fig.height = 17, fig.cap = "\\label{fig:figs}Anomalies in annual mean temperature, relative to 1950-2008 state average"}
change.barplot(temp_state, ylabel = "Temperature change (°F)", col_number = 5)
```







