---
title: "ModularityHW_Kelly"
author: "Michelle Kelly"
date: "June 21, 2018"
header-includes: 
- \usepackage{float} 
- \floatplacement{figure}{H} 
output: 
  pdf_document:
    fig_caption: yes
bibliography: Modularity_Citations.bib
---  
<!-- MODULARITY LESSON 8: have defined purpose for each Rmd chunk-->
```{r setup, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
# Global options
knitr::opts_chunk$set(fig.height = 5, fig.width = 15, cache=TRUE) 
# Enable caching in all chunks, Set default size of figures to 5x9

# Packages
library(dplyr)
library(data.table)
library(ggplot2)
library(maps)
library(maptools)
library(mapdata)
library(viridis)
library(cowplot)
library(grid)
library(gridExtra)
```

```{r dataImport, echo=FALSE}
# Dependencies: 
#     USAAnnualPcpn1950_2008.rds
#     USAAnnualTemp1950_2008.rds

provided <- TRUE # TRUE if data loaded from file, FALSE if data pulled from API

if (provided == TRUE){
  precip <- readRDS(file = "USAAnnualPcpn1950_2008.rds") # annual precipitation data for USA
  temp <- readRDS(file = "USAAnnualTemp1950_2008.rds") # annual temperature data for USA
}
```

```{r dataCleanup, echo=FALSE}
# MODULARITY LESSON 1: variables, not constants
minYears <- 40 # minimum years of data for each station

# MODULARITY LESSON 7: Pseudocode for trimming function

# trim <- function(dataset, minimumTimespan){
#   exclude years that have no data
#   count number of years of data for each station
#   if (years of data <= 40){
#     exclude station from dataframe
#   }
#   return(dataframe)
# }

# MODULARITY LESSON 2: write extensible code by creating trimming function

cleanByYear <- function(dataframe, nYears){
  dataframe <- na.exclude(dataframe)
  dataframe$count <- table(dataframe$name)[dataframe$name]
  for (row in 1:nrow(dataframe)){
    if (dataframe$count[row] <= nYears){
      dataframe[row,] <- NA
    }
  }
  dataframe <- na.exclude(dataframe)
  dataframe$count <- NULL
  return(dataframe)
}

precip <- cleanByYear(precip, minYears)
temp <- cleanByYear(temp, minYears)
```

# Introduction  

Average surface temperature on Earth has increased by 0.7&deg;C since the onset of the industrial era [@UnitedNationsFrameworkConventiononClimateChangeUNFCC2007]. Widespread fossil fuel use, and the following spike in atmospheric CO2, have acted as a global insulation blanket: trapping heat in the atmosphere and slowing it's release [@Asfaw2018]. Increased temperatures are also impacting precipitation regimes; in the United States, increases in the severity of drought and flood events are expected [@Dore2005]. However, the impact of global change on a local level is expected to be more spatially heterogenous, with different regions of the US experiencing varied climate and precipitation responses [@Trenberth2011].  

We assess these expectations using `r minYears` years of weather station data, asking (1) can climate warming be seen across the United States, (2) are the effects of climate warming regionally-dependent, and (3) what change has occured in precipitation?

# Methods  

We used annualy averaged weather-station temperature (`r length(unique(temp$state))` states, `r length(unique(temp$name))` weather stations) and precipitation data (`r length(unique(precip$state))` states, `r length(unique(precip$name))` weather stations) from `r min(temp$year)` to `r max(temp$year)`. Only weather stations with at least `r minYears` years of measurement data were included in this analysis.

Yearly summary statistics were calculated by taking the mean observation within a state during any given year. To compare changes across time, a mean value for the duration of observations within a state was calculated, and subtracted from the annual value (Equation \ref{eq-dev}).
  
\begin{equation}
\label{eq-dev}
Deviation\, from\, average_{state} = Annual\, average_{state} - Time\, series\, average_{state}
\end{equation}

Therefore, negative values represent time periods when a state was experiencing below average temperatures, and positive values represent periods when a state was experiencing above average temperatures.  

# Results and Discussion  

```{r dataAnalysis, echo = FALSE}
# yearlyStatistics: mean data across USA for each year
yearlyStatistics <- function(dataframe){
  dataframe %>% group_by(year) %>% summarize(mean = mean(data))
}

precip_summary <- yearlyStatistics(precip)
temp_summary <- yearlyStatistics(temp)

# stateStatistics: mean data across each state for each year

# stateStatistics pseudocode:
#   1. compute overall average, so we can compare values to the average
#   2. group the data by year and state
#   3. compute the mean, standard deviation, and difference of values from the yearly average
#       where negative difference values represent colder/less rain than average
#       positive values represent warmer/more rain than average

stateStatistics <- function(dataframe){
  dataframe_state <- 
    dataframe %>% 
      group_by(year, state) %>% 
      summarize(mean = mean(data))
  
  dataframe_statesummary <- 
    dataframe_state %>%
      group_by(state) %>%
      summarise(state_mean = mean(mean))
  
  dataframe_merge <- 
    merge(dataframe_state, dataframe_statesummary, by = "state")
}

precip_state <- stateStatistics(precip)
temp_state <- stateStatistics(temp)

# deviation: compute difference between state yearly average and state overall (1950-2008) average
deviation <- function(dataset_state){
  dataset_state$mean - dataset_state$state_mean
}

precip_state$change <- deviation(precip_state)
temp_state$change <- deviation(temp_state)

# timeseries.change: compute mean deviation from state overall average across 1950-2008
timeseries.change <- function(dataset_state){
  dataset_state %>%
    group_by(state) %>%
    summarize(sum.change = sum(change))
}

precip.timeseries.change <- timeseries.change(precip_state)
temp.timeseries.change <- timeseries.change(temp_state)
```

```{r dataVis_functions, echo = FALSE}
# heatmap: visualization of change through time using colored tiles
heatmap <- function(dataset, legendlabel, ...){
  ggplot(dataset, aes(x = state, y = year, ...)) +
  geom_tile(aes(fill = change), color = "white") +
  scale_fill_distiller(type = "div", palette = 7, name = legendlabel) +
  xlab("State") + ylab("Year") +
  theme_classic() +
  theme(legend.position = "bottom")
}

# change.barplot: visualization of change through time using bars
change.barplot <- function(dataset_state, ylabel, col_number, ...){
  ggplot(dataset_state, aes(x = year, y = change, ...)) +
    geom_col(aes(fill = change)) +
    facet_wrap(~ state, ncol = col_number) +
    scale_fill_distiller(type = "div", palette = 7) +
    xlab("Year") + ylab(ylabel) +
    theme_classic() +
    theme(panel.background = element_rect(color = "black"),
          legend.position = "none")
}
```

```{r significance_testing, echo = FALSE, warning = FALSE, message = FALSE}
# significance testing: yearly means
sigtest_precip <- lm(precip_summary$mean ~ precip_summary$year)

sigtest_temp <- lm(temp_summary$mean ~ temp_summary$year)

# significance testing: change by state
temp_lm <- summary(lm(temp_state$change ~ temp_state$year)) # signif
temp_lm_pval <- temp_lm$coefficients

precip_lm <- summary(lm(precip_state$change ~ precip_state$year)) #signif
precip_lm_pval <- precip_lm$coefficients
```

We see that deviations in temperature from local averages have generally increased from 1950-2008 (p-value < 0), with rapid increases from 1980 onwards (Figure \ref{fig:mapping}). However, warming has not been homogenous across the USA: states at northern latitudes and those located in the American Southwest have experienced more extreme effects.

We know that the Earth has undergone two main periods of warming within the last century, one from 1910 and 1945, and the other from 1976 to present day [@Walther2002]. These results support the latter half of that assessment. It is also known that the Southwest and Upper Midwest regions have experienced the highest increases in annual temperature above normal since 1976 [@Walther2002].  

\pagebreak 

```{r mapping, echo = FALSE, fig.height = 6, fig.cap="\\label{fig:mapping}Change in temperature per decade, relative to 1950-2008 state average."}
# data for mapping: plot mean deviation from average over last century of data (1998-2008) compared to mean deviation from average during 1950-1960
temp.change.map <- function(dataset_state, minYear, maxYear, axis_lims, legendPos){
  dataset_state <- 
    dataset_state %>%
      group_by(state, year) %>%
      filter(year >= minYear & year <= maxYear) %>%
      group_by(state) %>%
      summarize(mean.change = mean(change)) 
  
  states <- map_data("state") # pull state data from maps package
  states$region <- toupper(states$region) # change character data to uppercase
  setnames(states, "region", "state") # change column name
  
  dataset_state$state <- toupper(state.name[match(dataset_state$state, state.abb)])
  
  dataset_state <- inner_join(states, dataset_state, by = "state") # join dataframes
  
  # output map
  ggplot(data = dataset_state) +
    geom_polygon(aes(x = long, y = lat, group = group, fill = mean.change), 
                 color = "white") +
    coord_fixed(1.3) +
    scale_fill_distiller(limits = axis_lims, type = "div", palette = 7,
                         name = "Temperature 
change
relative to
average (°F)
") +
    theme_classic() +
    ggtitle(paste(minYear, maxYear, sep = "-")) +
    theme(legend.position = legendPos,
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
}

grid.arrange(
  temp.change.map(temp_state, minYear = 1950, maxYear = 1960, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 1960, maxYear = 1970, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 1970, maxYear = 1980, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 1980, maxYear = 1990, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 1990, maxYear = 2000, 
                  axis_lims = c(-2,2), legendPos = "none"),
  temp.change.map(temp_state, minYear = 2000, maxYear = 2008, 
                  axis_lims = c(-2,2), legendPos = "none"),
  get_legend(temp.change.map(temp_state, minYear = 1950, maxYear = 1960, 
                             axis_lims = c(-2,2), legendPos = "right")),
  layout_matrix = rbind(c(1,2,3,7),
                        c(4,5,6,7)),
  widths = c(3,3,3,1), heights = c(2,2)
)

```

Precipitation dynamics have changed from `r min(temp$year)` to `r max(temp$year)` (p-value = `r precip_lm_pval[8]`). Visually, we can see that there are both intense "high" and "low" years (that is, a potential increase in flood and drought events) (Figure \ref{fig:heatmap_precip}). To further investigate this phenomenon, we could use United States Geological Survey (USGS) river discharge data as an additional flood and drought predictor. It is not clear by this analysis alone whether location is a predictor of precipitation change. A finer scale (county level) analysis may be useful.  

```{r heatmap_precip, echo=FALSE, fig.height = 5, fig.cap="\\label{fig:heatmap_precip}Anomalies in annual precipitation, relative to 1950-2008 state average"}
# Anomalies in annual precipitation, relative to 1950-2008 state average
grid.arrange(
  heatmap(precip_state, legendlabel = "Precipitation 
change (in)"),
  change.barplot(precip_state, ylabel = "Precipitation change (in)", col_number = 3),
  layout_matrix = rbind(c(1,2,2,2),
                        c(1,2,2,2))
)
```

In summary, we have determined that climate warming from `r min(temp$year)` to `r max(temp$year)` is highly visually apparent. Generally, the southwest and upper midwest regions are experiencing the most warming, while the southern and southwest regions experienced colder than average temperatures during 1960-1980. We suspect that precipitation has become more "dynamic", with more flood and drought events, although further investigation is needed before a definitive conclusion.

\pagebreak 

# Supplemental
<!-- MODULARITY LESSON 3: Rewriting/revising your code can be good and important, and should not depress you, see unused plot here and many many versions on git -->
```{r barplot_temp, echo = FALSE, fig.height = 18, fig.cap = "\\label{fig:barplot_temp}Anomalies in annual mean temperature, relative to 1950-2008 state average"}
change.barplot(temp_state, ylabel = "Temperature change (°F)", col_number = 5)
```

\pagebreak 

# References
